\begin{thebibliography}{99}

\bibitem{Clarke} Edmund M.~Clarke,~Jr., Orna Grumberg and Doron A.~Peled. Model
  Checking. MIT Press, 1999.
\bibitem{Velder} Вельдер С. Э., Шалыто А. А. Введение в верификацию автоматных
  программ. XXXVI научная и учебно-методическая конференция
  профессорско-преподавательского и научного состава СПбГУ ИТМО, 2007.
\bibitem{Dijkstra} Dijkstra, E. W. Hierarchical ordering of sequential processes. , Acta
  Informatica 1(2), June~1971, pp.~115-138.
\bibitem{SpinRoot} Model checker SPIN. \url{http://spinroot.com/}.
\bibitem{BitHash1} Gerard J. Holzman. An Analysis of Bit-state
  Hashing. In Proc. 15th Int. Conf on. Protocol Specification,
  Testing, and Verification, pp. 301–314.
\bibitem{StateCompr} Gerard J. Holzmann. State Compression in SPIN: Recursive
  Indexing and Compression Training Runs. In Proceedings of Third International
  SPIN Workshop, 1997.
\bibitem{POD} Gerard J. Holzmann. The Verification of Concurrent Systems,
  unpublished manuscript. AT\&T Inc., 1995, Chapter 4.
\bibitem{DLTL1} Jiri Barnat, Lubos Brim. Distributed LTL
  Model-Checking in SPIN. Lecture Notes in Computer Science,
  vol. 2057/2001, pp 200-216.
\bibitem{DLTL2} Jiri Barnat, Lubos Brim. Parallel Breadth-First Search
  LTL Model-Checking. 18th IEEE International Conference on Automated
  Software Engineering (ASE'03), pp. 106-115.
\bibitem{LT99} Flavio Lerda, Riccardo Sisto. Distributed-Memory Model
  Checking with SPIN. Lecture Notes In Computer Science, vol. 1680, pp
  22-39.
\end{thebibliography}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "main"
%%% End: 
